---
title: "Python"
editor: source
## ------  general Output Options
execute:     
  eval: false
  echo: true     #  include source code in output
  warning: false  #  include warning code in output
  error: false    #  include error code in output
  output: false   # include output code in output (CHG in BLOCKS)
  # include: false   # R still runs but code and results DON"T appear in output  
  cache: false # normalmnte false
toc: true
fig-cap-location: top
tbl-cap-location: top
format:
  html:
    # theme: flatly #spacelab
    code-fold: false # redundant bc echo false 
    toc-depth: 3
    toc_float: true
    toc-location: left
    toc-title: Outline
    embed-resources: true # external dependencies embedded (Not in ..._files/)
  # pdf:
  #   toc-depth: 2
  #   toc-title: Indice
  #   highlight-style: github
  #   #lang: it
  #   embed-resources: true # external dependencies embedded (Not in ..._files/)
format-links: false
---
 
RStudio has `reticulate` package embedded

Python is standalone in your machine
```{bash}
python3 --version
#Python 3.13.0
```

```{r}
print("Hello World")
```


```{python}
print("Hello World")
```

```{python}
print("Hello World" *3)
```

```{python}
x = 5
x
list = [1, 2, 3, "Hello"]
list
# utile per creare liste di cose
my_dict = {"name": "John", "age": 30, "city": "New York"}
print(my_dict)
```


```{r}
add_1 <- function(x) {
  x + 1
}
```

```{python}
def add_1(x):
    return x + 1 # must be 4 spaces 
   
```

```{r}
reticulate::py_module_available("numpy")       # Returns TRUE if available
reticulate::py_module_available("pandas")      # Check another package
```

# To install a package 

### --- (Install ARM-native brew if needed)
This is because Apple Silicon Macs should use the ARM-native Homebrew, located at: /opt/homebrew/bin/brew

```{bash}
# /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
# echo >> ~/.zprofile
# echo 'eval "$(/opt/homebrew/bin/brew shellenv)"' >> ~/.zprofile
# eval "$(/opt/homebrew/bin/brew shellenv)"
```
Then check
```{bash}
which brew
# should return      /opt/homebrew/bin/brew OK!
```

But I still have Python and other things that Brew installed in usr/local/bin (WRONG in this MAC)
```{bash}
ls -l /usr/local/bin/python3*

# if any
sudo rm -f /usr/local/bin/python3.13 \
             /usr/local/bin/python3.13-config \
             /usr/local/bin/python3-config \
             /usr/local/bin/python3.bak
# then 
#no matches found 
```


### --- (Install Python 3.11 via ARM Homebrew  if needed)

> This should now install an ARM-native Python to /opt/homebrew/bin/python3.
(Instead for some reason it tells me that it is installing to /usr/local/bin/python3.11, which is wrong for my machine.)

```{bash}
/opt/homebrew/bin/brew reinstall python@3.11
#==> Summary
#🍺  /opt/homebrew/Cellar/python@3.11/3.11.12: 3,306 files, 62.1MB

which python3 # STILL returns /usr/bin/python3
# this is because the symlink python3 → python3.11 was not created by Homebrew (by design), and your shell doesn’t know to default to the version you installed via ARM Homebrew.

# BUT here it gives both 
file $(which python3) # it must say arm64 not x86_64
# (MINE both )            
# /usr/bin/python3 (for architecture x86_64):     Mach-O 64-bit executable x86_64
# /usr/bin/python3 (for architecture arm64e):     Mach-O 64-bit executable arm64e

# (IT SHOULD BE)   /opt/homebrew/bin/python3: Mach-O 64-bit executable arm64
```


### --- Link python3.11 into /opt/homwbrew/bin
You've got Python 3.11 installed via ARM Homebrew, but it's not symlinked into your main PATH at /opt/homebrew/bin. That's why /opt/homebrew/bin/python3 doesn't exist, even though the files live at: `/opt/homebrew/Cellar/python@3.11/3.11.12/...` 

...to fix with an alias for python3
```{bash}
# add an alias to your ~/.zprofile
file.edit("~/.zshrc")
# add this line at the bottom
# alias python3=/opt/homebrew/bin/python3.11

# then
source ~/.zshrc

# and check again 
which python3 
# python3: aliased to /opt/homebrew/bin/python3.11
file /opt/homebrew/bin/python3.11
# /opt/homebrew/bin/python3.11: Mach-O 64-bit executable arm64   OK 
```

### --- 1. [Create] Use a virtual environment for reticulate

> OKKIO:
+ On Intel Macs, Homebrew installs to: /usr/local/
+ On Apple Silicon Macs (like yours), Homebrew installs to: /opt/homebrew/
(This avoids mixing Intel and ARM binaries.)

 
### --- 2. Activate the environment (optional for shell)

Shows that it is using `arm64`, which is the correct architecture for your machine.

```{r}
reticulate::conda_list()
#          name                                                                   python
#1         base                   /Users/luisamimmi/Library/r-miniconda-arm64/bin/python
#2 r-reticulate /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/bin/python

```
 
### --- 3. RESTART RStudio

### --- 4.a ⛔️ Make THE VIRTUAL ENVIRONMENT permanent (just once):
After restarting RStudio, you need to tell RStudio to use the virtual environment you just created. You can do this in several ways:

```{r}
#reticulate::use_virtualenv("~/.venvs/r-reticulate", required = TRUE)
```

### --- 4.b ✅ Make THE VIRTUAL ENVIRONMENT permanent (just once):
You're using Miniconda installed via reticulate, and there's already a conda environment called r-reticulate.
```{r}

reticulate::use_condaenv("r-reticulate", required = TRUE)

# Check if it works  !!! YES 
reticulate::py_config()
# python:         /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/bin/python
# libpython:      /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/lib/libpython3.10.dylib
# pythonhome:     /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate:/Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticula# te
# version:        3.10.17 | packaged by conda-forge | (main, Apr 10 2025, 22:23:34) [Clang 18.1.8 ]
# numpy:          /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/lib/python3.10/site-packages/numpy
# numpy_version:  2.2.4
# 
# NOTE: Python version was forced by use_python() function
```

### --- 6. Add this to either your project-level .Rprofile or your global ~/.Rprofile:
```{r}
# 1 OPEN 
# Global Rprofile
file.edit("~/.Rprofile")

# 2 ADD this at the bottom 
# if (requireNamespace("reticulate", quietly = TRUE)) {
#   reticulate::use_condaenv("r-reticulate", required = TRUE)
# }
```

### --- 7. Check if it works
```{r}
reticulate::py_config()

# python:         /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/bin/python
# libpython:      /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/lib/libpython3.10.dylib
# pythonhome:     /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate:/Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticula# te
# version:        3.10.17 | packaged by conda-forge | (main, Apr 10 2025, 22:23:34) [Clang 18.1.8 ]
# numpy:          /Users/luisamimmi/Library/r-miniconda-arm64/envs/r-reticulate/lib/python3.10/site-packages/numpy
# numpy_version:  2.2.4
# 
# NOTE: Python version was forced by use_python() function

```


### --- 8. Check if it works

```{python}
import numpy as np
np.arange(5)
```

### --- [JUST IN CASE ]

If for any reason the conda environment doesn't load automatically, you can do so with:'
```{r}
reticulate::use_condaenv("r-reticulate", required = TRUE)
```

