---
title: "1.a) Tidyeval and functional programming"
date: "`r Sys.time()`"
date-format: "ddd, D of MMMM, YYYY"
date-modified: last-modified
lang: en
editor: source
engine: knitr
## ------  general Output Options
execute:     
  eval: false    # actually run? 
  echo: true     #  include source code in output
  warning: false  #  include warning code in output
  error: false    #  include error code in output
  output: false   # include output code in output (CHG in BLOCKS)
  # include: false   # R still runs but code and results DON"T appear in output  
  cache: false # normalmnte false
toc: true
fig-cap-location: top
tbl-cap-location: top
format:
  html:
    # theme: flatly #spacelab
    code-fold: false # redundant bc echo false 
    toc-depth: 3
    toc_float: true
    toc-location: left
    toc-title: Outline
    embed-resources: true # external dependencies embedded (Not in ..._files/)
  # pdf:
  #   toc-depth: 2
  #   toc-title: Indice
  #   highlight-style: github
  #   #lang: it
  #   embed-resources: true # external dependencies embedded (Not in ..._files/)
format-links: false
---

::: {.callout-caution collapse="false"}
Web page construction in progress...
:::



```{r}
library(tidyverse)
library(rlang)
```

# Learning resources

+ [ALbert Rapp YT vid](https://www.youtube.com/watch?v=L_sX-sL9aWM&t=1810s)
+ [Bruno Rodriguez YT vid](https://www.youtube.com/watch?v=pcvWKVlRmwE)
+ [rlang DATAMASKING](https://rlang.r-lib.org/reference/topic-data-		mask.html)
+ [Tidy eval in R: A simple example](https://www.youtube.com/watch?v=YlZuT6PWOEY&t=1s)
+ [Lionel Henry | Interactivity and Programming in the Tidyverse](https://www.youtube.com/watch?v=SW_TUkPXSmA) | RStudioConf (2020)


# Explanation
Premise: `tidyverse` functions use **tidy evaluation** (= *they don't evaluate the value of a variable right away*! = **Non-Standard evaluation**). 

  + (+) This means you can do some intermediate transformation to the variable in abstract (e.g. to a generic "column" thing)
  + (-) it’s hard to refer to variables indirectly, and hence harder to program with

In contrast, normal/base/custom R functions DO evaluate objects (i.e. `a+b`) **as soon as possible**   =  **Standard evaluation**  

So, to take full advantage of **Non-Standard evaluation** (more interactivity, but also writing custom functions), I will need a sort of **METAVARIABLE** (a "quosure"), i.e. something that doesn't get evaluated until I tell so.

## NON STANDARD EVALUATION in TIDYVERSE

+ **DEFUSING** (**DELAYING**) function arguments: I can create a "quosure" with `rlang::enquo()` /  `rlang::enquos()` so an expression can be examined, modified, and injected into other expressions.


## TWO (complementary) FORMS of NSE used in the Tidyverse 

### 1) **TIDY SELECTION** is used in SELECTION VERBS 
			
e.g. in `dplyr::select()` `across()`, `relocate()`, `rename()` and `pull()` use  ***tidy selection*** where expressions are either interpreted in the context of the data frame (e.g. `c(cyl, am)` or evaluated in the user environment (e.g. `all_of()`, `starts_with()`)
			
```{r}
#| eval: false

# EXE `across()`
summarise_mean <- function(data, vars) {
	# all variables selected by user...	
	data %>% summarise(n = n(), across({{ vars }}, mean))
}
# call 
starwars %>% 
	group_by(homeworld) %>% 
	# with where
	summarise_mean(where(is.numeric))
```
			
### 2) **DATA MASKING** used in ACTION VERBS 

ACTION VERBS  = `dplyr::mutate()`, `ggplot2::aes()`, `arrange()`, `count()`, `filter()`, `group_by()`,   and `summarise()`.

 Normal interactive programming (`tidyverse`) use  ***data-masking***, which allow you to use variables in the “current” data frame without any extra syntax. This:
 
 + (+) makes it nicer to interactively work (no extra typing of `data$column`, just `column`), but 
 + (-) makes it harder to create your own functions (it could be ambiguous what is a data-variable and what is an  env-variable).  

```{r}
#| eval: false

## ---  tidyverse non std eval 
starwars %>% 
 filter (homeworld == "Tatooine")  


```

SOOOOO We need some way to add `$` back into the picture. Passing data-masked arguments to functions requires **INJECTION** (= **quasiquotation**), i.e. **TO INJECT A FUNCTION ARGUMENT IN A DATA-MASKED CONTEST, YOU EMBRACE IT WITH** `{{`

Inside data-masking function (actions), we can use injection operators:

		+ `{{` embracing operator (`rlang`)
		+ `!!` operator (`base`) 
		+ `.data` pronouns. 
		+ `.env` pronouns.

## Defuse with "embracing" `{{variable }}` (inside custom f)  

```{r actv_embrace}

## --- own function with tidyverse non std eval ACTION VERB `group_by()`
my_summarise <- function(data, group_var) {
  data %>%
# tell to inject whatever argument supplied to the function (homeworld) 
		# **in place** of "group_var" 
    group_by({{ group_var }}) %>%
    summarise(mean = mean(height))
}

# call(s) 
my_summarise (starwars, homeworld) 
my_summarise (starwars, sex) 
```

 
:::{.callout-warning}
Without `{{group_var}}` I would get the error  
	*"! Must group by variables found in `.data`.
	✖ Column `group_var` is not found."*
:::


# Different options 


## 1   Defuse (nothing!) + Inject `{{`  (inside custom f) 

```{r}
# -------- OR 
grouped_mean_1 <- function(df, group_var, summarize_var) {
	df %>% 
# Defuse and inject in a single step with the embracing operator
    group_by({{group_var}} ) %>% 
    summarize(mean = mean({{summarize_var}} , na.rm = TRUE))
}

# call
grouped_mean_1(
  df = starwars, 
  group_var = sex, 
  summarize_var =  height 
)
```

## 2  Defuse `enquo`  + Inject `!!`  (inside custom f) 
```{r}
# We can tell group_by() not to quote by using !! (pronounced “bang bang”). !! says something like “evaluate me!” or “unquote!”test 

grouped_mean_2 <- function(df, group_var, summarize_var) {
	## -- Defuse the user expression in `*_var`
	group_var = enquo(group_var)
	summarize_var = enquo(summarize_var)

  df %>% 
  	## -- Inject the expression contained in `*_var` 
    group_by(!!group_var) %>% 
    summarize(mean = mean(!!summarize_var, na.rm = TRUE))
}

# call
grouped_mean_2(
  df = starwars, 
  group_var = sex, 
  summarize_var =  height 
)
```



## 3  Defuse `...`  +  Inject `...` 

In this case, `summarize_var` goes in front and `...` last 

+ `...` can stand for multiple variables 
```{r}
# ---- func 
grouped_mean_3 <- function(df,  summarize_var, ...) {
  
	## -- Defuse the summarize_var = enquo(summarize_var) 
	## ... group_var >>>> NO NEED FOR ENQUO with ... !
	summarize_var = enquo(summarize_var)
	
	df %>% 
		group_by(...) %>% 
		summarize(mean = mean(!!summarize_var, na.rm = TRUE))
}

# ---- call
grouped_mean_3(
	df = starwars, 
	sex, homeworld, # (...)
	summarize_var =  height 
)
```


:::{.callout-tip}
`{{...}}` Basically we are saying "everything I throw at the function will be carried along until I want to evaluate it"   
:::


# Different options (with left side eq)

## 1b (nothing!) + `{{` & left side `:=`  !!!!

+ Super compact left side syntax with `"sometext_{{group_var}}" :=`
```{r}
# --- func
grouped_mean_1b <- function(df, group_var, summarize_var) {
	df %>% 
		# Defuse and inject in a single step with the embracing operator
		group_by({{group_var}} ) %>% 
		summarize( "BY_{{group_var}}" := mean({{summarize_var}} , na.rm = TRUE))
}

# --- call
grouped_mean_1b (
	df = starwars, 
	group_var = sex, 
	summarize_var =  height 
)
```


## 2b `enquo` +  `!!` & left side `:=` 

2 things needed here :

	+  `as_label(enquo(____var))`
	+  left side syntax with `!!str_c("Mean_", ____var) :=`
 
```{r}
# --- func
grouped_mean_2b <- function(df, group_var, summarize_var) {
	## -- Defuse the user expression in `*_var`
	group_var = enquo(group_var)
	summarize_var = as_label(enquo(summarize_var)) # as_label(enquo !!!!!
	
	df %>% 
		## -- Inject the expression contained in `*_var` 
		group_by(!!group_var) %>% 
		summarize(!!str_c("Mean_", summarize_var) := mean(!!summarize_var, na.rm = TRUE))
}

# --- call
grouped_mean_2b(df = starwars, 
					 group_var = sex, 
					 summarize_var =  height 
)
```

## 3b `...` + `...` & left side `:=` 
```{r }
## --  define function 
grouped_mean_3b <- function(df,  summarize_var, ...) {
#	group_var = ...  NO NEED FOR ENQUO!
	summarize_var = enquo(summarize_var)
	summarize_var_name <- as_label(enquo(summarize_var))
	
	df %>% 
		group_by(...) %>% 
	#	summarize(!!summarize_var_name := mean(!!summarize_var, na.rm = TRUE))
	# or  
	summarize(!!str_c("My_mean_", summarize_var_name) := mean(!!summarize_var, na.rm = TRUE))
	# ERRORE ?!?!?!?
	# summarize(str_c("Mean_", !!summarize_var_name) := mean(!!summarize_var, na.rm = TRUE))
}

## --  call function
grouped_mean_3b(df = starwars, 
					 sex, homeworld, #	group_var
					 summarize_var =  height 
)
```

:::{.callout-warning}
OKKIO!!! 
Strange enough... seems like the unquoting must be of the WHOLE left-side of the equation not just of the quoted variable as I thought 
	+ `!!summarize_var_name :=  ... ` OK 
	+ `!!str_c("Mean_", summarize_var_name) :=  ... ` OK: xchè??????
	+ `str_c("Mean_", !!summarize_var_name) :=  ... ` WRONG: xchè??????

:::


# Using  `.data`
It’s good practice to prefix *named arguments* with a . (`.data`)to reduce the risk of conflicts between your arguments and the arguments passed to `...`
```{r}
## --  define function 
grouped_mean_4 <- function(data,  summarize_var, ...) {
  
#	group_var = enquo(group_var) NO NEED FOR ENQUO!
	summarize_var = enquo(summarize_var)
	summarize_var_name <- as_label(enquo(summarize_var))

	data %>% 
		group_by(...) %>% 
	#	summarize(!!summarize_var_name := mean(!!summarize_var, na.rm = TRUE))
	# or even better 
	summarize(!!str_c("Mean_", summarize_var_name) := mean(!!summarize_var, na.rm = TRUE))
	
}

## --  call function
grouped_mean_4(
	data = starwars, 
	summarize_var =  height ,
	sex, homeworld
)
```


# Examples

```{r}
 
starwars <- starwars

# I FORWARD a (masked)  argument with DOUBLE EMBRACE 
my_summarise <- function(data, var) {
  data %>% dplyr::summarise(Mean =mean( {{ var }}, na.rm = TRUE ))
}

## --  call function
call <- my_summarise (starwars, height)

# The .data pronoun is a tidy eval feature that is enabled in all data-masked arguments, just like {{
my_summarise2 <- function(data, var) {
  data %>% dplyr::summarise(mean = mean(.data[[var]], na.rm = TRUE ))
}

call2 <- my_summarise2 (starwars, "height")



```

```{r}

# # ------- 1/3 FUNZIONE: raggruppo (2 gruppi) e calcolo qualche proporzione 
# # https://www.youtube.com/watch?v=pcvWKVlRmwE
# 
# f_prop_grouping <- function(start_df, end_df, group_var1, group_var2, dummy1, dummy2  ) {
#   
#   end_df <- start_df %>% 
#     # grouping var(s)
#     group_by( {{group_var1}} ,  {{group_var2}} # misura e stato
#     ) %>% 
#     summarise(n_group = n(),# n_stato
#               TOT_dummy1 = sum({{dummy1}}, na.rm = TRUE ), # SUM_dummy1
#               TOT_dummy2 = sum({{dummy2}}, na.rm = TRUE ) # SUM_dummy2
#     ) %>% 
#     mutate(N_group = sum(n_group), # N_stato
#            Perc_group =  paste0(round(n_group/N_group,3)*100 ,"%"), # %_stato
#            Perc_group_dummy1 = paste0(round(TOT_dummy1/n_group,3)*100 ,"%"), # % dummy over n(group)
#            Perc_group_dummy2 = paste0(round(TOT_dummy2/n_group,3)*100 ,"%"),
#            
#     ) }

# run
# end_df <- f_prop_grouping(AL_anagr_stato_t, end_df, misura, stato, inizio_past, inizio_rit )

# ------- 2/3 FUNZIONE: raggruppo (2 gruppi) e calcolo qualche proporzione 
f_prop_grouping2 <- function(start_df, end_df, group_var1, group_var2, dummy1, dummy2 ) {  
  # This to use the "walrus operator" := on the LEFT (naming the derived vars )
  dummy1name <- as_label(enquo(dummy1))
  dummy2name <- as_label(enquo(dummy2))
 
  
  end_df <- start_df %>% 
    # grouping var(s)
    group_by( {{group_var1}}, {{group_var2}} # misura e stato
    ) %>% 
    summarise(n_group = n(),# n_stato
              !!str_c("TOT_", dummy1name) := sum({{dummy1}}, na.rm = TRUE ), # SUM_dummy1
              !!str_c("PERC_", dummy1name) := paste0(round(sum({{dummy1}}, na.rm = TRUE)/n_group,3)*100 ,"%"), # % dummy over n(group)
              
              !!str_c("TOT_", dummy2name) := sum({{dummy2}}, na.rm = TRUE ),  # SUM_dummy2
              !!str_c("PERC_", dummy2name) := paste0(round(sum({{dummy2}}, na.rm = TRUE)/n_group,3)*100 ,"%")   # % dummy over n(group) 
               ) %>% 
    # ? NOT SURE WHY HERE IT ONLY CONSIDERS 1ST GROUPING VARIABLE
    mutate (N_group = sum(n_group),
            Perc_group =  paste0(round(n_group/N_group,3)*100 ,"%")  # %_stato
    ) %>% 
    relocate (c("N_group","Perc_group" ), .before = !!str_c("TOT_", dummy1name))  
}

# # ------- 3/3 FUNZIONE: raggruppo (2 gruppi) e calcolo qualche proporzione 
# # using [“enquo” + “!!”  ]      |  "syms" function and the “!!!”  (for multiple vars) 
# f_prop_grouping3 <- function(start_df, end_df, group_vars, dummy1, dummy2  ) {
#   # define the list of group_by vars "syms" 
#   group_vars <- syms(group_vars)
#   # This to use the "walrus operator" := on the LEFT (naming the created var  
#   dummy1name <- as_label(enquo(dummy1))
#   dummy2name <- as_label(enquo(dummy2))
#   
#   end_df <- start_df %>% 
#     # call grouping var(s) “!!!” 
#     group_by( !!!group_vars ) %>%    
#     summarise(n_group = n(),# n_stato
#               # TOT_dummy1 = sum({{dummy1}}, na.rm = TRUE ), # SUM_dummy1
#               # TOT_dummy2 = sum({{dummy2}}, na.rm = TRUE ) # SUM_dummy2
#               !!str_c("TOT_", dummy1name) := sum({{dummy1}}, na.rm = TRUE ), # SUM_dummy1
#               !!str_c("PERC_", dummy1name) := paste0(round(sum({{dummy1}}, na.rm = TRUE)/n_group,3)*100 ,"%"), # % dummy over n(group)
#               
#               !!str_c("TOT_", dummy2name) := sum({{dummy2}}, na.rm = TRUE ),  # SUM_dummy2
#               !!str_c("PERC_", dummy2name) := paste0(round(sum({{dummy2}}, na.rm = TRUE)/n_group,3)*100 ,"%") # % dummy over n(group) 
#               ) %>% 
#     # ? NOT SURE WHY HERE IT ONLY CONSIDERS 1ST GROUPING VARIABLE
#     mutate (N_group = sum(n_group),
#             Perc_group =  paste0(round(n_group/N_group,3)*100 ,"%")  # %_stato
#     ) %>% 
#     relocate (c("N_group","Perc_group" ), .before = !!str_c("TOT_", dummy1name))  
#     
#   
# }


# ------- 2/3 FUNZIONE: raggruppo (2 gruppi) e calcolo qualche MEDIA
f_mean_grouping2 <- function(start_df, end_df, group_var1, group_var2, numer1, numer2) {  
  # This to use the "walrus operator" := on the LEFT (naming the derived vars )
  numer1name <- as_label(enquo(numer1))
  numer2name <- as_label(enquo(numer2))
  
  # operations
  end_df <- start_df %>% 
    # grouping var(s)
    group_by( {{group_var1}}, {{group_var2}} # misura e stato
    ) %>% 
    summarise(n_group = n(),# n_stato
              !!str_c("Media_", numer1name) := mean({{numer1}}, na.rm = TRUE),
              !!str_c("Media_", numer2name) := mean({{numer2}}, na.rm = TRUE)
              ) %>% 
    # ? NOT SURE WHY HERE IT ONLY CONSIDERS 1ST GROUPING VARIABLE
    mutate (N_group = sum(n_group),
            Perc_group =  paste0(round(n_group/N_group,3)*100 ,"%")  # %_stato
    ) %>% 
    relocate (c("N_group","Perc_group" ), .before = !!str_c("Media_", numer1name) )
}

# OKKIO!!!!!! 
#end_df <- f_prop_grouping2(AL_anagr_stato_t, end_df, c("misura", "stato"), inizio_past, inizio_rit )


# f_prop_grouping2 <- function(start_df, end_df, group_var1, group_var2, ... ) {
#   
#   end_df <- start_df %>% 
#     # grouping var(s)
#     group_by( {{group_var1}} ,  {{group_var2}} # misura e stato
#     ) %>% 
#     summarise(n_group = n(),# n_stato
#               TOT_dummy1 = sum(..., na.rm = TRUE ), # SUM_dummy1
#               TOT_dummy2 = sum(..., na.rm = TRUE ) # SUM_dummy2
#     ) %>% 
#     mutate(N_group = sum(n_group), # N_stato
#            Perc_group =  paste0(round(n_group/N_group,3)*100 ,"%"), # %_stato
#            Perc_group_dummy1 = paste0(round(TOT_dummy1/n_group,3)*100 ,"%"), # % dummy over n(group)
#            Perc_group_dummy2 = paste0(round(TOT_dummy2/n_group,3)*100 ,"%"),
#            
#     ) }



# ------- FUNZIONE: generalizzo il nome della fase (prefix) 
f_rimuovo_pref <- function(data, prefix = "word_"){
  rename_with(.data = data, 
              .cols = dplyr::starts_with(prefix), # (default e' everything() e le pescava comunque)
              # rinomino le date eliminando il prefisso
              .fn = function(x)sub(prefix,"",x))
  
}

# EXE uso
#BC_PROGETTAZIONE_temp <- f_rimuovo_pref(BC_PROGETTAZIONE , prefix = "PROG_ESEC_") 


# -------- FUNZIONE: introduco qualche calcolo sulle date delle fasi procedurali
f_calcoli_date <- function(data, inizio_fase_prev, inizio_fase_eff, fine_fase_prev, fine_fase_eff) {
  dplyr::mutate(data,
                durata_prev =  {{fine_fase_prev}} - {{inizio_fase_prev}} ,
                #durata_eff =   {{fine_eff}} - {{inizio_eff}} ,
                inizio_V_today = case_when(
                  {{inizio_fase_prev}} <= today() ~ "pre_oggi",
                  {{inizio_fase_prev}} > today() ~  "post_oggi",
                  TRUE ~ "Ignoto"),
                inizio_discrep = {{inizio_fase_eff}} - {{inizio_fase_prev}} ,
                inizio_ritardo = case_when(
                  inizio_V_today == "Ignoto" ~ "[Fase non prevista]",
                  inizio_V_today == "post_oggi" ~ "Inizio previsto futuro",
                  inizio_V_today == "pre_oggi" & !is.na({{inizio_fase_eff}}) ~ if_else(
                    inizio_discrep > 0 , glue("rit = {inizio_discrep} gg"), glue("ant = {inizio_discrep} gg")
                  ),
                  inizio_V_today == "pre_oggi" & is.na({{inizio_fase_eff}}) ~  "No inizio effettivo")
  )
}

# EXE uso
# BC_PROGETTAZIONE_calc <-  f_calcoli_date (
#   data = BC_PROGETTAZIONE_temp , 
#   inizio_fase_prev, inizio_fase_eff, fine_fase_prev, fine_fase_eff)

```


# MORE

https://jonthegeek.com/2018/06/04/writing-custom-tidyverse-functions/
